---
title: "EAS 648 - Lab 1"
author: "Abas Shkembi"
date: "2023-10-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F)
library(tidyverse)
library(sf)
```

# Questions

# Question 1

*Discuss the advantages and challenges associated with an open data science approach. Provide an example based on this week’s reading. (1-2 paragraphs)*

**Response:** 

### Reading in the data

```{r, warning=F, message=F}
df_boulder <- st_read("Data/boulder_data_eas648/BoulderSocialMedia.shp")
```

The plot below examines how the points from the dataset are spatially distributed. The `alpha` argument was set to `0.01` to visualize where most of the pictures are being taken.

```{r}
ggplot() +
    geom_sf(data = df_boulder,
    fill = NA, alpha = .01, shape = 21) +
    theme_bw()
```

### Investigating where the points are concentrated the most

Below, we visualize where most of the pictures are being taken using a 2D density plot. We can see that most of the pictures taken are concentrated between $105.28^\circ W$ and $105.29^\circ W$ longitudinally, and between $39.99^\circ N$ and $40.00^\circ N$ latiduinally.

```{r}

# transform the crs of the dataframe
df_boulder <- st_transform(df_boulder, crs = "+proj=longlat +ellps=WGS84 +datum=WGS84")

# create dummy dataframe
points_pic <- df_boulder

# convert simple features to data frame
points <- as.data.frame(points_pic) #convert to dataframe
points$lon <- unlist(map(points_pic$geometry,1)) #adding lon column
points$lat <- unlist(map(points_pic$geometry,2)) #adding lat column

ggplot(points, aes(x = lon, y = lat)) +
  geom_density2d_filled() +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10))

```

I reclassifed the points that fall within this boundary as being "high density" (labeled as `high_dens`). The figure below visualizes this classifiation.

```{r}
points %>%
  mutate(high_dens = 
           ifelse(
             (lon > -105.29 & lon < -105.28) & (lat > 39.99 & lat < 40.00),
             1,
             0
           )) %>%
  ggplot() +
  geom_point(aes(x = lon, y = lat, color = as.character(high_dens)), alpha = .01, shape = 21) +
  theme_bw()
```

I then chose to employ a gradient boosted classification tree model to identify which characteristics are most indicative of a picture being taken at a "high density" area of the park. I used all of the distance variables (e.g., distance to nearest climbing wall) and the elevation variable as predictors of an area being "high density" vs "not high density". The results of the model may elucidate why the "high density" area of the park has so many pictures being taken there, and what characteristics of the park may be making so many people take a picture in this area.

```{r}
# load in library for gradient boosted regression tree model
library(gbm)

# select variables of interst
df_gbm <- points %>%
  mutate(high_dens = 
           ifelse(
             (lon > -105.29 & lon < -105.28) & (lat > 39.99 & lat < 40.00),
             1,
             0
           ))  %>%
  select(high_dens, 4:12) %>% # our outcome and the distance variables
  na.omit() %>% as_tibble()

# run model
set.seed(4234); boost_dens <- gbm(high_dens ~ ., 
                          data = df_gbm, 
                          distribution = 'bernoulli', # our outcome is 1 vs 0
                          cv.folds = 5,
                          shrinkage = 0.01,
                          verbose = FALSE,
                          n.trees = 1000, 
                          interaction.depth = 5)

summary(boost_dens, plotit = FALSE) %>%
  as_tibble() %>%
  mutate("Variable Importance" = paste0(round(rel.inf, 1), "%")) %>%
  select(-2) %>%
  DT::datatable()


```

The findings from the gradient boosted regression tree model suggest that a picture's distance from the prairie dog mounds and a picture's distance from hiking trails are the two most important indicators of a picture being taken in a "high density" area. Let's examine the individual relationships of each indicator with the outcome.

```{r}

# get variable names 
variables_gbm <- df_gbm %>% select(-1) %>% colnames()

# create variable name crosswalk
var_crosswalk <- tibble(
  variable = variables_gbm,
  nice_var_name = c("Dist. climbing wall", "Dist. hiking trl", "Dist. nat. landmark",
                    "Dist. walking trl", "Dist. biking trl", "Dist. prairie dog", 
                    "Elevation", "Dist. lakes, river", "Dist. street")
)

# get order of variables by variable importance from the model
factor_levels <- summary(boost_dens, plotit = FALSE) %>% as_tibble() %>% .$var

# sort the variable crosswalk in order of most important to least important variable
var_crosswalk <- var_crosswalk %>% mutate(variable = factor(variable, levels = factor_levels)) %>% arrange(variable)


```

```{r}

# create function to extract odds ratios for each variable
OR_gbm <- function(var) {
  
  boost_dens %>%
  pdp::partial(plot=FALSE, n.trees = 1000, pred.var = var) %>% # calculate partial dependency
  as_tibble() %>% # convert to tibble
  mutate(variable = var) %>% # rename inputted column to "variable"
  rename("value" = 1) %>% # rename the first column to "value"
  mutate(yhat = 1/(1+exp(-yhat))) %>% # calculate probability
  mutate(yhat = yhat/(1-yhat)) %>% # calculate odds
  mutate(median = median(df_gbm[[var]])) %>% # find median value of variable
  mutate(median_yhat = yhat[which.min(abs(value - median))]) %>% # find OR associated with median of variable
  mutate(OR = yhat/median_yhat) %>% # calculate odds ratio relative to median odds
  select(variable, value, OR) # select the variable name, value, and odds ratio
  
}

# create single data frame of odds ratios for each variable
OR_df <- NULL
for(i in 1:length(variables_gbm)) {
  temp_df <- OR_gbm(variables_gbm[i])
  
  OR_df <- rbind(OR_df, temp_df)
}


OR_df %>%
  left_join(var_crosswalk, by = "variable") %>%
  mutate(nice_var_name = factor(nice_var_name, levels = var_crosswalk$nice_var_name)) %>%
  ggplot(aes(x = value, y = OR)) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  facet_wrap(~nice_var_name, scales = "free_x") +
  labs(x = "Distance (in ft.)", y = "Odds ratio", title = "Odds of a picture being taken in high density area of park") +
  theme_bw()


```

The figure demonstrates that generally the closer people are to the prairie dog mounds and hiking trails, the more likely they are to take a picture. These two factors play the largest role in why people are taking a picture in the "high density" area of the park.


